Efficiently summarizing relationships in large samples: a general duality between statistics of genealogies and genomes

### 2019-09-16 treestats_paper.tex
Editing pass.

Before formal definitions, we will give an overview of the general framework.
Roughly speaking, to compute a statistic we will
propagate ``weights'' additively up each tree,
summarize these weights on each branch with a ``summary function'',
and aggregate these summaries to produce statistics.
The abstract idea of summarizing weights propagated up each tree
is only a small generalization of a familiar operations in population genetics:
summarizing the allele frequency spectrum.
For instance, we could calculate how many of a certain set $A$ of samples
inherit from a particular branch in a tree
by (a) assigning each of these samples weight 1 (and other nodes weight zero), then
(b) finding the total weight of all nodes in the subtree inheriting from that branch.
Any mutation falling on this branch would be inherited by
This gives the number of samples that would inherit any mutation falling on this branch,
so is analogous to the allele frequency spectrum.
Figure~\ref{fig:divergence_diagram} shows a simple example of these procedures
of propagating weights up a tree and summarizing them,
used to calculate mean sequence divergence over a short sequence spanned by a single tree.

% Many population genetic summary statistics can be defined using the allele frequency spectrum of a collection of homologous sequences.
If we only considered summaries of the allele frequency spectrum of a single sample,
our statistics would be equivalent to that described by \citet{fu1995statistical}.
To connect the framework to familiar notions,
here is a reframing of \citeauthor{fu1995statistical}'s statistics.
Write $p_i(a)$ for the frequency of allele $a$ at a site $i$,
and $h(p)$ for the allele frequency spectrum
--- i.e., the number of alleles whose frequency is $p$ in our sample ---
and suppose that for some function $f()$, we want to compute $\sum_p h(p) f(p)$.
(For instance, mean genetic diversity is computed with $f(p) = p (1-p)$.)
Since this is a simple sum over sites,
we can compute this by summing over the $L$ loci, as
\begin{align*}
    \sum_p h(p) f(p) = \sum_{i=1}^L \sum_a f(p_i(a)).
\end{align*}
(Note that since the ancestral allele is included, this function is invariant to relabeling of alleles.)
If we know where each mutation has occurred on the genealogy,
we can easily find the frequency of the corresponding allele by counting how many samples
inherit from that mutation in the tree (and accounting for those that subsequently mutate).

@@

We allow vector-valued weights,
i.e., $\iw(v)$ may be a vector $(\iw_1(v), \ldots, \iw_m(v))$,
so that when summarizing we have access to more than one aspect of each subtree.

@@

We will use weights that count number of samples inheriting from each node
sufficiently often that we give them a special name ---
for a subset set $S$ of samples,
the \emph{indicator weights of $S$} are the sample weights $\bone_S$ with
$\bone_S(u) = 1$ if $u \in S$ and $\bone_S(u) = 0$ otherwise.

@@

@@

\begin{figure}
    \includegraphics{{figures/swept.1000.999.1e-09.diversity}.pdf}
    \caption{
        Mean genetic diversity and time to most recent common ancestor
        in 500Kb windows along a 50Mb genome (0.5 Morgans) following several selective sweeps.
        In each case, ``site'' is mean genetic diversity (Tajima's $\pi$) divided by mutation rate,
        and ``branch'' is the corresponding branch statistic described above.
        The tree sequence was produced by simulating mutations under positive selection with mutation rate $10^{-12}$
        in a population of size 1000 for 400 generations using SLiM,
        followed by recapitation with $N_e=1000$~\citep{haller2018tree}.
        The selected alleles at the marked sites have selection coefficients between 0.08 and 0.25,
        and are at frequencies 96.8\%, 100\%, 16.25\%, 100\%, and 82.6\% in the final generation, respectively.
        All curves use the same tree sequence, including selected mutations,
        and with additional neutral mutations added.
        \textbf{(top)}
        Diversity within the entire population,
        computed using for mode ``site'' from 20 independent assignments of mutations to the same tree sequence with mutation rate $\mu = 10^{-9}$.
        \textbf{(middle)}
        As in the top panel, but with mutation rate $\mu=10^{-8}$,
        showing that as mutation rate increases, mode ``site'' (divided by $\mu$) converges to mode ``branch''.
        \textbf{(bottom)}
        ``Site'' and ``branch'' diversity within 20 disjoint samples of size 100 each,
        computed on a single tree sequence with mutation rate $\mu = 10^{-9}$.
        \label{fig:sweep_duality}
    }
\end{figure}

@@

We do not know the true genealogies underlying real data,
but recent methods are available to estimate them at scale.
In Figure~\ref{fig:sweep_duality}, we showed that branch and site statistics
matched well in simulated data. However, these simulations make many
simplifying assumptions, and, moreover,
the underlying tree topologies and branch lengths are exactly
correct, which no inference method can hope to achieve.
To evaluate the correspondence between site and branch statistics
in tree sequences inferred from real data,
we calculated mean sequence diversity along chromosome 20 from the 1000 Genomes dataset
in 1Mb windows separately in each of the five continental groupings,
calculated using the site statistic described in Example~\ref{ex:site_diversity},
and compared this to the dual branch statistic (of Example~\ref{ex:branch_diversity})
from the trees and node times estimated by Relate~\citep{speidel2019method}.
All calculations were done only using the portions of the chromosome
passing the 1000 Genomes Phase 1 ``strict'' mask for sequencing accessibility.
% from ftp://ftp.1000genomes.ebi.ac.uk/vol1/ftp/phase1/analysis_results/supporting/accessible_genome_masks/README.20120515_phase1_pop_gen_mask
As shown in Figure~\ref{fig:chr20}, Branch and Site diversity along the genome are very similar.
Does this mean that Relate is doing a better job at inferring the tree sequence
Is this an indication that Relate is doing a good job at inferring trees,
or just a byproduct of the inference method (which, naturally, estimates branch lengths to be proportional to numbers of mutations)?
Answering this question requires a deeper understanding of the processes that shape
genetic diversity along the genome.

@@

In practice, we do not expect branch and site diversity to agree exactly,
because of local differences in mutation rate and intensities of selected mutations.
For instance, regions in which a high proportion of potential mutations are deleterious
are expected to have a lower diversity for two reasons:
first, the deleterious mutations themselves are less likely to be found,
and second, the indirect effect of deleterious mutations on nearby sites reduces typical tree height
and thus diversity at even neutral, linked sites \citep{hudson1994levels,charlesworth1997effects}.
The first effect would cause site and branch statistics to differ,
because it effectively reduces the mutation rate in the region
and violates the assumption of independence of mutations given the trees.
However, the second effect does \emph{not} affect the correspondence between site and branch statistics,
since it is mediated by tree shape.
It is generally unknown how much mutation rate varies along the genome
\citep[but see][]{supek2015differential}
or how dense are targets of selection \citep{leffler2012revisiting}.
For this reason, it is very interesting to see how close a correspondence between site and branch
statistics it is \emph{possible} to obtain --
it is tempting to say that the best tree sequence would obtain
as tight a match between site and branch statistics as possible,
with remaining variation explained by direct selection and/or mutation rate variation.

@@

%%%%%%%%%%%%%%%%%%%%%
\subsection*{Algorithms}
In this section we describe and analyse the algorithm used
to compute branch length statistics. This is a generalization
of the algorithm used to maintain the number of samples from a
given set that derive from each node in a tree
sequence \cite[Algorithm L]{kelleher2016efficient}. The central
problem shared across site, branch and node statistics
is to efficiently maintain the ``state'' $x_u$ (where $x_u$ is the
sum of the weights for all samples descending
from, and including, node $u$; see Figure~\ref{fig:divergence_diagram}).
As we move along the sequence, the trees change when branches are inserted and removed.
By carefully ordering these insertions and removals,
we ensure that this state can be correctly maintained
using only small adjustments for each tree.
Computing the statistics is then a straightforward matter of applying the summary function $f$
to the node states and aggregating the results appropriately for the
particular statistics mode and windowing options.

@@

\algstep{B2.}{Terminate.}{If $j \geq |\Et|$ or $t_l \geq L$,
    return $\sigma / L$ and terminate.
}

\algstep{B3.}{Edge removal loop.}{If $k \geq |\Et|$ or $t_l \neq
    \Et[\indexout_k]\prop{right}$ go to \algref{B6}.}

@@

\algstep{B6.}{Edge insertion loop.}{If $j \geq |\Et|$ or $t_l \neq
    \Et[\indexin_j]\prop{left}$ go to \algref{B9}.}

@@

Algorithm \algref{B} begins by setting the initial state for $\pi$, $\beta$,
$x$ and $F$ for each node in the tree sequence, and then sets the values
of $x_u$ and $F_u$ for each of the samples $u$ in $S$.
(The initial state is the ``empty forest'', where no nodes are connected to any others.)
Afterwards, we set our ``running sum'' $s$ and output statistic $\sigma$ to zero, along
with the ``tree left'' variable $t_l$. The $j$ and $k$ variables are
used to keep track of our position in the edge insertion and edge
removal indexes, respectively. After completing initialisation in
\algref{B1}, we enter the main tree loop in \algref{B2} which is
run once for each tree in the sequence. As we are processing each
tree, we keep track of the edges that need to be processed with
the $t_l$ variable, which stores the left-most genome coordinate
in the current tree. The first thing we do for a new tree is to
remove any edges that were in the previous tree are not in the
current tree. These must be the edges in which the right coordinate
is equal to $t_l$, and so \algref{B3} loops over these edges using
the edge removal index $\indexout$. Step \algref{B4} then removes
the branch $u \mapsto v$ corresponding to a single edge, by
subtracting its contribution to the running sum $s$, setting the parent of
$u$ to $-1$ and its branch length to $0$. As shown in
Figure~\ref{fig:algorithm_example}B, removing an edge will result
in the subtree rooted at $u$ become disconnected from the rest of the
tree. Step \algref{B5} ensures that the state of the rest of the tree
is correctly maintained by propagating the loss of the state $x_u$
up the tree from $v$. Thus, for each node $v$ that was a
direct ancestor of $u$ in the previous tree, we first subtract
its contribution from the running sum $s$ and remove the contribution
of $x_u$ from its state. Since the state of $v$ has changed, we must
recompute the value of our summary cache $F_v$, and then finally
add the new contribution of $v$ to $s$. (For example, note the changes in $x$
and $F$ for nodes 7 and 8 in Figure~\ref{fig:algorithm_example}B.)

Once all of the edges with right coordinate equal to $t_l$ have been removed in
steps \algref{B3--B5}, we then insert edges that start in the current tree
(i.e., with left coordinate $t_l$) in steps \algref{B6--B8}. We update the
state to account for the new branch $u \mapsto v$ by setting the parent of
$u$ to $v$, computing the branch length of $u$ and adding the new contribution
of $u$ to $s$ in \algref{B7}. Then, as adding a new edge can connect the
subtree rooted at $u$ to the larger tree at $v$, we propagate the gain of $x_u$
up the tree in step \algref{B8} (note the symmetry with step \algref{B5}).
Finally, once we have removed and inserted all of the relevant edges, we are
ready to add the contribution of the current tree to the overall statistic,
$\sigma$ and move on to the next tree in step \algref{B9}. To do this, we first
compute the right-hand coordinate of the tree $t_r$ (a process slightly
complicated by the possible presence of ``gaps'' in the tree sequence, spanned
by no edges). Then, we add the running sum $s$ weighted by the span of the
current tree $t_r - t_l$ to $\sigma$, and return to \algref{B2} to process the
next tree. If we have reached the end of the sequence, we then divide $\sigma$
by the sequence length to renormalise and exit.

To analyse Algorithm~\algref{B} we will assume that the tree sequence has
been simulated under the standard coalescent with a sample of size
$n$ and population scaled recombination rate $\rho = 4 N_e r$,
where $r$ is the mean number of recombinations per chromosome per unit of time.
Under these conditions, there are $O(n + \rho \log n)$ edges in the tree sequence
\citep{kelleher2016efficient}.  Clearly, each edge is examined
exactly once in steps \algref{B6} and \algref{B7} in order to create
the trees, and at most once in steps \algref{B3} and \algref{B4} (we do
not remove the edges for the final tree). Additionally, each edge that
we insert or remove after the initial building of the tree
may incur the cost of traversing up the tree as
far as the root in steps \algref{B5} and \algref{B8}.
As coalescent genealogies are asymptotically
balanced \citep{li2013coalescent}, the expected number of nodes
on the path to root is $\log_2 n$.
Therefore, the overall running time
of Algorithms~\algref{B} is of order $n + \rho \log(n) \log_2(n)$, which is
\begin{equation}\label{eqn:alg_complexity}
    O\left(n + \rho \log(n)^2 \right)
\end{equation}
Computing a site statistic will have the same complexity,
as the same internal state is maintained,
and the number of segregating sites is proportional to the number of edges
(with constant equal to the ratio of mutation to recombination rate).
The asymptotic analysis here assumes that we traverse upwards to root
for each edge, but this is not the case. Edges are removed in
oldest-first order, guaranteeing that the deepest node any subtree being
modified is processed first. Therefore, this subtree will be disconnected from the
main tree, ensuring that we traverse upwards all the way to root only once.
Similarly, edges are inserted in youngest-first order, ensuring that subtrees are only
reconnected to the main tree when they are fully built.

@@

We used coalescent simulations to compare the performance of calculating Tajima's \citeyearpar{Tajima1989-de} $D$
statistic from tree sequences to calculating it from integer matrices containing genotypes at all variable positions.
\autoref{fig:stats_performance} shows that the tree sequence methods implemented in \texttt{tskit} processes variant data
substantially faster than matrix-based approaches once the sample size is above $n \approx 500$ haploids --
three times faster for one thousand haplotypes,
growing to nearly three orders of magnitude faster for one million haplotypes.
The expected number of mutations for each replicate is $10^4 \times \sum_{i=1}^{n-1}\frac{1}{i}$, and thus ranges from
$\approx 32,000$ to $\approx 138,000$ \citep{Watterson1975-ej}.
\autoref{fig:stats_performance} only considers the time required to calculate the statistic,
and does not include the time required to build the genotype matrix, nor the memory required to store it.
For the largest samples size of $n=10^6$,
the matrix is approximately 200 gigabytes, and thus not practical on many systems.
The largest tree sequence simulated required less than one gigabyte of memory.

@@

% speed/efficiency: big data sets; large simulations
The most obvious application of these methods on current practice
is to improve efficiency of existing pipelines,
as tree sequences allow storage and processing of large genomic data sets
with orders of magnitude less space and time than most other methods.
All statistics described here (and more) are implemented
in the fast, rigorously tested library \tskit{}
that also provides tools for working with tree sequences in Python and C.
The largest efficiency gains are found for very large sample sizes
only likely to be available in real data sets for humans in the near future,
but the ability to analyze millions of whole genomes
is useful for understanding evolutionary simulations of any species
\citep[see][for a recent example]{galloway2019stickleback}.

% duality: site stats ~ branch stat, but conversely branch more accurate version of site
%  maybe analogous to imputation
We hope a more broadly interesting and useful aspect of this paper
is a more in-depth look at the correspondence between statistics calculated with genome sequence
and properties of the underlying genealogies, i.e.,
the duality between Site and Branch statistics under the infinite-sites model.
This general correspondence is well-known,
but the common framework implemented here makes it easy to translate between the two.
One way to use this translation is to answer:
what aspect of tree shape is this population genetics statistic summarizing?
This can help when interpreting results, especially if reality may not fit an idealised model
of separate populations.
However, methods to infer tree sequences from real data
make it now possible to work in the other direction:
instead of calculating (Site) statistics from the genotype matrix,
we can calculate precisely analogous (Branch) statistics from the trees themselves,
thus hopefully bypassing the extra layer of noise induced by mutation.
How well this works will depend on the ability of inference methods
to estimate the true tree sequence.
This might plausibly produce less noisy estimates because tree sequence inference
essentially uses the signal from nearby patterns of variation
to interpret relationships at a given site,
thus transforming the simple binary split induced by a SNP
into a much richer source of information at every variant site.
Furthermore,
if tree sequence inference can be made insensitive to local variation in mutation rate,
calculation of branch statistics would provide a summary method
that is not affected by this potentially important confounding factor.

@@

%%%%%%%%%%%%%%%%%%%%%%%
\subsection*{Acknowledgments}
Thanks to Graham Coop, Andrew Kern, Nick Patterson, and Kelley Harris for useful suggestions
and to Wilder Wohns, Yan Wong and Leo Speidel for providing inferred tree sequences.

>>>>

Before formal definitions, we will give an overview of the general framework.
Roughly speaking, to compute a statistic we will
propagate ``weights'' additively up each tree,
summarize these weights on each branch with a ``summary function'',
and aggregate these summaries to produce statistics.
The abstract idea of summarizing weights propagated up each tree
is only a small generalization of a familiar operations in population genetics:
summarizing the allele frequency spectrum.
For instance, we could calculate how many of a certain set $A$ of samples
inherit from a particular branch in a tree
by (a) assigning each of these samples weight 1 (and other samples weight zero), then
(b) finding the total weight of all samples in the subtree inheriting from that branch.
This gives the number of samples that would inherit any mutation falling on this branch,
so is analogous to the allele frequency spectrum.
Figure~\ref{fig:divergence_diagram} shows a simple example of these procedures
of propagating weights up a tree and summarizing them,
used to calculate mean sequence divergence over a short sequence spanned by a single tree.

% Many population genetic summary statistics can be defined using the allele frequency spectrum of a collection of homologous sequences.
If we only considered summaries of the allele frequency spectrum of a single sample,
our statistics would be equivalent to that described by \citet{fu1995statistical}.
To connect the framework to familiar notions,
here is a reframing of \citeauthor{fu1995statistical}'s statistics.
Write $p_i(a)$ for the frequency of allele $a$ at a site $i$,
and $h(p)$ for the allele frequency spectrum
--- i.e., the number of alleles whose frequency is $p$ in our sample ---
and suppose that for some function $f()$, we want to compute $\sum_p h(p) f(p)$.
(For instance, mean genetic diversity is computed with $f(p) = p (1-p)$.)
Since this is a simple sum over sites,
we can compute this by summing over the $L$ loci, as
\begin{align*}
    \sum_p h(p) f(p) = \sum_{i=1}^L \sum_a f(p_i(a)).
\end{align*}
% (Note that since the ancestral allele is included, this function is invariant to relabeling of alleles.)
If we know where each mutation has occurred on the genealogy,
we can easily find the frequency of the corresponding allele by counting how many samples
inherit from that mutation in the tree (and accounting for those that subsequently mutate).

@@

We allow vector-valued weights,
i.e., $\iw(v)$ may be a vector $(\iw_1(v), \ldots, \iw_m(v))$,
so that when summarizing we have access to more than one aspect of each subtree.
We will often be interested in statistics defined in terms of different
subsets of our samples (corresponding to, e.g., different populations); it
is useful to have some additional notation to define these statistics.
Given a subset set $S$ of samples,
the \emph{indicator weights of $S$} are the sample weights $\bone_S$ with
$\bone_S(u) = 1$ if $u \in S$ and $\bone_S(u) = 0$ otherwise.

@@

@@

\jk{What does this last sentence mean?}

@@

\begin{figure}
\centering
    \includegraphics[width=0.8\textwidth]{{figures/swept.1000.999.1e-09.diversity}.pdf}
    \caption{
        Mean genetic diversity and time to most recent common ancestor
        in 500Kb windows along a 50Mb genome (0.5 Morgans) following several selective sweeps.
        In each case, ``site'' is mean genetic diversity (Tajima's $\pi$) divided by mutation rate,
        and ``branch'' is the corresponding branch statistic described above.
        The tree sequence was produced by simulating mutations under positive selection with mutation rate $10^{-12}$
        in a population of size 1000 for 400 generations using SLiM,
        followed by recapitation with $N_e=1000$~\citep{haller2018tree}.
        The selected alleles at the marked sites have selection coefficients between 0.08 and 0.25,
        and are at frequencies 96.8\%, 100\%, 16.25\%, 100\%, and 82.6\% in the final generation, respectively.
        All curves use the same tree sequence, including selected mutations,
        and with additional neutral mutations added.
        \textbf{(top)}
        Diversity within the entire population,
        computed using for mode ``site'' from 20 independent assignments of mutations to the same tree sequence with mutation rate $\mu = 10^{-9}$.
        \textbf{(middle)}
        As in the top panel, but with mutation rate $\mu=10^{-8}$,
        showing that as mutation rate increases, mode ``site'' (divided by $\mu$) converges to mode ``branch''.
        \textbf{(bottom)}
        ``Site'' and ``branch'' diversity within 20 disjoint samples of size 100 each,
        computed on a single tree sequence with mutation rate $\mu = 10^{-9}$.
        \label{fig:sweep_duality}
    }
\end{figure}

@@

We do not know the true genealogies underlying real data,
but recent methods are available to estimate them at
scale~\citep{harris2019database,kelleher2019inferring,speidel2019method}.
In Figure~\ref{fig:sweep_duality}, we showed that branch and site statistics
matched well in simulated data. However, these simulations make many
simplifying assumptions, and, moreover,
the underlying tree topologies and branch lengths are exactly
correct (which inference methods can only hope to approximate).
To evaluate the correspondence between site and branch statistics
in tree sequences inferred from real data,
we calculated statistics for chromosome 20
of the 1000 Genomes dataset~\citep{1k2015global} using dated trees estimated
by Relate~\citep{speidel2019method}.
Specifically, we calculated diveristy using the site statistic
described in Example~\ref{ex:site_diversity},
and compared this to the dual branch statistic (Example~\ref{ex:branch_diversity})
in 1Mb windows in each of the five continental groupings.
All calculations were done only using the portions of the chromosome
passing the 1000 Genomes Phase 1 ``strict'' mask for sequencing accessibility.
% from ftp://ftp.1000genomes.ebi.ac.uk/vol1/ftp/phase1/analysis_results/supporting/accessible_genome_masks/README.20120515_phase1_pop_gen_mask
As shown in Figure~\ref{fig:chr20}, Branch and Site diversity along the genome are very similar.
Is this an indication that Relate is doing a good job at inferring trees,
or just a byproduct of the inference method (which, naturally, estimates branch lengths to be proportional to numbers of mutations)?
Answering this question requires a deeper understanding of the processes that shape
genetic diversity along the genome.

@@

In practice, we do not expect branch and site diversity to agree exactly,
because of local differences in mutation rate and intensities of selected mutations.
For instance, regions in which a high proportion of potential mutations are deleterious
are expected to have a lower diversity for two reasons:
first, the deleterious mutations themselves are less likely to be found,
and second, the indirect effect of deleterious mutations on nearby sites reduces typical tree height
and thus diversity at even neutral, linked sites \citep{hudson1994levels,charlesworth1997effects}.
The first effect would cause site and branch statistics to differ,
because it effectively reduces the mutation rate in the region
and violates the assumption of independence of mutations given the trees.
However, the second effect does \emph{not} affect the correspondence between site and branch statistics,
since it is mediated by tree shape.
It is generally unknown how much mutation rate varies along the genome
\citep[but see][]{supek2015differential}
or how dense are targets of selection \citep{leffler2012revisiting}.
For this reason, it is very interesting to see how close a correspondence between site and branch
statistics it is \emph{possible} to obtain ---
it is tempting to say that the best tree sequence would obtain
as tight a match between site and branch statistics as possible,
with remaining variation explained by direct selection and/or mutation rate variation.

@@

%%%%%%%%%%%%%%%%%%%%%
\subsection*{Algorithms}
In this section we describe and analyse the algorithm used
to compute Branch statistics. This is a generalization
of the algorithm used to maintain the number of samples from a
given set that derive from each node in a tree
sequence \cite[Algorithm L]{kelleher2016efficient}. The central
problem shared across site, branch and node statistics
is to efficiently maintain the ``state'' $x_u$ (where $x_u$ is the
sum of the weights for all samples descending
from, and including, node $u$; see Figure~\ref{fig:divergence_diagram}).
As we move along the sequence, the trees change when branches are inserted and removed.
By carefully ordering these insertions and removals,
we ensure that this state can be correctly maintained
using only small adjustments for each tree.
Computing the statistics is then a straightforward matter of applying the summary function $f$
to the node states and aggregating the results appropriately for the
particular statistics mode and windowing options.

@@

\algstep{B2.}{Terminate.}{If $j = |\Et|$ return $\sigma / L$ and terminate.
}

\algstep{B3.}{Edge removal loop.}{If $k = |\Et|$ or $t_l \neq
    \Et[\indexout_k]\prop{right}$ go to \algref{B6}.}

@@

\algstep{B6.}{Edge insertion loop.}{If $j = |\Et|$ or $t_l \neq
    \Et[\indexin_j]\prop{left}$ go to \algref{B9}.}

@@

Algorithm \algref{B} begins by setting the initial state for $\pi$, $\beta$,
$x$ and $F$ for each node in the tree sequence, and then sets the values
of $x_u$ and $F_u$ for each of the samples $u$ in $S$.
(The initial state is the ``empty forest'', where no nodes are connected to any others.)
Afterwards, we set our ``running sum'' $s$ and output statistic $\sigma$ to zero, along
with the ``tree left'' variable $t_l$. The $j$ and $k$ variables are
used to keep track of our position in the edge insertion and edge
removal indexes, respectively. After completing initialisation in
\algref{B1}, we enter the main tree loop in \algref{B2} which is
run once for each tree in the sequence. As we are processing each
tree, we keep track of the edges that need to be processed with
the $t_l$ variable, which stores the left-most genome coordinate
in the current tree. The first thing we do for a new tree is to
remove any edges that were in the previous and tree are not in the
current tree. These must be the edges in which the right coordinate
is equal to $t_l$, and so \algref{B3} loops over these edges using
the edge removal index $\indexout$. Step \algref{B4} then removes
the branch $u \mapsto v$ corresponding to a single edge, by
subtracting its contribution to the running sum $s$, setting the parent of
$u$ to $-1$ and its branch length to $0$. As shown in
Figure~\ref{fig:algorithm_example}B, removing an edge will result
in the subtree rooted at $u$ become disconnected from the rest of the
tree. Step \algref{B5} ensures that the state of the rest of the tree
is correctly maintained by propagating the loss of the state $x_u$
up the tree from $v$. Thus, for each node $v$ that was a
direct ancestor of $u$ in the previous tree, we first subtract
its contribution from the running sum $s$ and remove the contribution
of $x_u$ from its state. Since the state of $v$ has changed, we must
recompute the value of our summary cache $F_v$, and then finally
add the new contribution of $v$ to $s$. (For example, note the changes in $x$
and $F$ for nodes 7 and 8 in Figure~\ref{fig:algorithm_example}B.)

Once all of the edges with right coordinate equal to $t_l$ have been removed in
steps \algref{B3--B5}, we then insert edges that start in the current tree
(i.e., with left coordinate $t_l$) in steps \algref{B6--B8}. We update the
state to account for the new branch $u \mapsto v$ by setting the parent of
$u$ to $v$, computing the branch length of $u$ and adding the new contribution
of $u$ to $s$ in \algref{B7}. Then, as adding a new edge can connect the
subtree rooted at $u$ to the larger tree at $v$, we propagate the gain of $x_u$
up the tree in step \algref{B8} (note the symmetry with step \algref{B5}).
Finally, once we have removed and inserted all of the relevant edges, we are
ready to add the contribution of the current tree to the overall statistic,
$\sigma$ and move on to the next tree in step \algref{B9}. To do this, we first
compute the right-hand coordinate of the tree $t_r$ (a process slightly
complicated by the possible presence of ``gaps'' in the tree sequence, spanned
by no edges). Then, we add the running sum $s$ weighted by the span of the
current tree $t_r - t_l$ to $\sigma$, and return to \algref{B2} to process the
next tree. If we have reached the end of the sequence, we then divide $\sigma$
by the sequence length to normalise and exit.

To analyse Algorithm~\algref{B} we will assume that the tree sequence has
been simulated under the standard coalescent with a sample of size
$n$ and population scaled recombination rate $\rho = 4 N_e r$,
where $r$ is the mean number of recombinations per chromosome per unit of time.
Under these conditions, there are $O(n + \rho \log n)$ edges in the tree sequence
\citep{kelleher2016efficient}.  Clearly, each edge is examined
exactly once in steps \algref{B6} and \algref{B7} in order to create
the trees, and at most once in steps \algref{B3} and \algref{B4} (we do
not remove the edges for the final tree). Additionally, each edge that
we insert or remove after the initial building of the tree
may incur the cost of traversing up the tree as
far as the root in steps \algref{B5} and \algref{B8}.
As coalescent genealogies are asymptotically
balanced \citep{li2013coalescent}, the expected number of nodes
on the path to root is $\log_2 n$.
Therefore, the overall running time
of Algorithms~\algref{B} is of order $n + \rho \log(n) \log_2(n)$, which is
\begin{equation}\label{eqn:alg_complexity}
    O\left(n + \rho \log(n)^2 \right).
\end{equation}
Computing a site statistic will have the same complexity,
as the same internal state is maintained,
and the number of segregating sites is proportional to the number of edges
(with constant equal to the ratio of mutation to recombination rate).
The asymptotic analysis here assumes that we traverse upwards to root
for each edge, but this is not the case. Edges are removed in
oldest-first order, guaranteeing that the deepest node any subtree being
modified is processed first. Therefore, this subtree will be disconnected from the
main tree, ensuring that we traverse upwards all the way to root only once.
Similarly, edges are inserted in youngest-first order, ensuring that subtrees are only
reconnected to the main tree when they are fully built.

@@

We used coalescent simulations to compare the performance of calculating Tajima's \citeyearpar{Tajima1989-de} $D$
statistic from tree sequences to calculating it from integer matrices containing genotypes at all variable positions.
\autoref{fig:stats_performance} shows that the tree sequence methods implemented in \texttt{tskit} processes variant data
substantially faster than matrix-based approaches once the sample size is above $n \approx 500$ haploids ---
three times faster for one thousand haplotypes,
growing to nearly three orders of magnitude faster for one million haplotypes.
The expected number of mutations for each replicate is $10^4 \times \sum_{i=1}^{n-1}\frac{1}{i}$, and thus ranges from
$\approx 32,000$ to $\approx 138,000$ \citep{Watterson1975-ej}.
\autoref{fig:stats_performance} only considers the time required to calculate the statistic
once the data is present in each library's native format.
For the largest samples size of $n=10^6$,
the matrix is approximately 200 gigabytes, and thus not practical on many systems.
The largest tree sequence simulated required less than one gigabyte of memory.

@@

% speed/efficiency: big data sets; large simulations
The most obvious application of these methods on current practice
is to improve efficiency of existing pipelines,
as tree sequences allow storage and processing of large genomic data sets
with orders of magnitude less space and time than most other methods.
All statistics described here (and more) are implemented
in the rigorously tested \tskit{} library,
which provides a suite tools for efficiently working with tree sequences in Python and C.
The largest efficiency gains are found for very large sample sizes
only likely to be available in real data sets for humans in the near future,
but the ability to analyze millions of whole genomes
is useful for understanding evolutionary simulations of any species
\citep[see, e.g.,][]{martin2017human,galloway2019stickleback}.

% duality: site stats ~ branch stat, but conversely branch more accurate version of site
%  maybe analogous to imputation
We hope a more broadly interesting and useful aspect of this paper
is a more in-depth look at the correspondence between statistics calculated with genome sequence
and properties of the underlying genealogies, i.e.,
the duality between Site and Branch statistics under the infinite-sites model.
This general correspondence is well-known,
but the common framework implemented here makes it easy to translate between the two.
One way to use this translation is to answer:
what aspect of tree shape is a particular population genetics statistic summarizing?
This can help when interpreting results, especially if reality may not fit an idealised model
of separate populations.
However, methods to infer tree sequences from real data
make it now possible to work in the other direction:
instead of calculating (Site) statistics from the genotype matrix,
we can calculate precisely analogous (Branch) statistics from the trees themselves,
thus hopefully bypassing the extra layer of noise induced by mutation.
How well this works will depend on the ability of inference methods
to estimate the true tree sequence.
This might plausibly produce less noisy estimates because tree sequence inference
may use the signal from nearby patterns of variation
to interpret relationships at a given site,
thus transforming the simple binary split induced by a SNP
into a much richer source of information at every variant site.
Furthermore,
if tree sequence inference can be made insensitive to local variation in mutation rate,
calculation of branch statistics would provide a summary method
that is not affected by this potentially important confounding factor.

@@

%%%%%%%%%%%%%%%%%%%%%%%
\subsection*{Acknowledgments}
Thanks to Graham Coop, Andrew Kern, Nick Patterson, and Kelley Harris for useful suggestions
and Leo Speidel for providing the inferred 1000 Genomes tree sequence.
